---
title: "Series-02-Rstudio : correlations and statistical tests"
author: "Delphine HUND"
date: "`r format(Sys.time(), '%d %B, %Y, %H:%M')`"
output:
  html_document:
    df_print: paged
    code_folding: hide
    toc: false
    number_sections: true
---

# Correlations and linear regression
In this series of exercises, you will explore correlations and linear regression, using a dataset from a real experiment.

## The problem
You want to know if there is a relationship between the nonuse of the proximal part of the upper limb (PANU) and the nonuse of the shoulder (SANU) or the elbow (EENU).

## The data
As you will see, the data is far from perfect, but this is the reality of experimental data, especially in the case of clinical data.

(Download the file NonUse.csv from the course Moodle.)

```{r load data}
nonuse <- read.csv("data/test_data/NonUse.csv", sep=",", header=TRUE)
head(nonuse)
summary(nonuse)
```

This file contains measures of upper extremity nonuse:

    PANU: Proximal Arm Non Use
    SANU: Shoulder Antepulsion Non Use
    EENU: Elbow Extension Non Use

In your notebook, make a clear description of the experiment, the variables and the data. Do not forget to give the units of each variable, and the range of possible values.

| name | description | unit | range
|---|-------|--|-----
| PANU | Proximal Arm Non Use | % | O-100 |
| SANU | Shoulder Antepulsion Non Use | % | O-100 |
| EENU | Elbow Extension Non Use | % | O-100 |

## Corelation analysis
Below are some questions to guide your analysis :

- Correlation between PANU/SANU
We used a Spearman correlation, because we don't have assumptions about the distribution of the data. We do not assume linear relationship between variables.
```{r correlation PANU-SANU}
cor.test(nonuse$PANU, nonuse$SANU, method = "spearman")
```
- Correlation between PANU/EENU
```{r correlation PANU-EENU}
cor.test(nonuse$PANU, nonuse$EENU, method = "spearman")
```

- PANU-SANU relationship

```{r plot SANU-lm}
library(ggplot2)
ggplot(nonuse, aes(x=PANU, y=SANU)) +
  geom_point() +
  geom_smooth(method='lm', formula= y~x, color='blue', se=T) +
  # xlim(-10, 60) +
  # ylim(-50, 50) +
  
labs(title="PANU vs. SANU", x="PANU (%)", y="SANU (%)")
```

- PANU-EENU relationship

```{r plot EENU-lm}

ggplot(nonuse, aes(x=PANU, y=EENU)) +
  geom_point() +
  geom_smooth(method='lm', formula= y~x, color='blue', se=T) +
  # xlim(-10, 60) +
  # ylim(-50, 50) +
  
labs(title="PANU vs. EENU", x="PANU (%)", y="EENU (%)")
theme_minimal()
theme_minimal()
```

- Linear regression for PANU-SANU

```{r lm/SANU}
lm_sanu <- lm(SANU ~ PANU, data=nonuse)
summary(lm_sanu)
coef(lm_sanu)
slope_1 <- coef(lm_sanu)[2]
intercept_1 <- coef(lm_sanu)[1] #?
cat("Regression equation: SANU=", slope_1, "PANU +", intercept_1, "\n")
```

- Linear regression for PANU-EENU

```{r lm/EENU}
lm_eenu <- lm(EENU ~ PANU, data=nonuse)
summary(lm_eenu)
coef(lm_eenu)
slope_2 <- coef(lm_eenu)[2]
intercept_2 <- coef(lm_eenu)[1] 
cat("Regression equation: EENU=", slope_2, "PANU +", intercept_2, "\n")
```
- Regression equations: PANU = a * SANU + b and PANU = a * EENU + b

```{r equations PANU}
a <- 1 / slope_1
b <- -intercept_1 / slope_1

cat("Regression equation : PANU =", a, "* SANU +", b, "\n")

# PANU = a*EENU + b
a <- 1 / slope_2
b <- -intercept_2 / slope_2

cat("Regression equation : PANU =", a, "* EENU +", b, "\n")
```

- Results

Spearman’s rank correlation rho showed a weak but significant positive association between PANU and SANU (ρ = 0.216, p = 0.0017), while a moderate and significant positive association was found between PANU and EENU (ρ = 0.387, p < 0.0001). Linear regression analyses confirmed these results: no significant relationship was observed between SANU and PANU (SANU = 0.089 × PANU + 0.586, R² = 0.009, p = 0.171), whereas a significant positive relationship was found between EENU and PANU (EENU = 0.565 × PANU + 3.284, R² = 0.225, p < 0.0001). When PANU was expressed as a function of SANU and EENU, the regression equations obtained were PANU = 11.195 × SANU – 6.561 and PANU = 1.771 × EENU – 5.816,  respectively.
(with GPT's help)

- What are the limits of your analysis?




