---
title: "Series-02-Rstudio : Data mining and statistical tests"
author: "Delphine HUND"
date: "`r format(Sys.time(), '%d %B, %Y, %H:%M')`"
output:
  html_document:
    df_print: paged
    code_folding: hide
    toc: false
    number_sections: true
---
# Correlations and linear regression

In this series, the aim is to explore correlations and linear regression, using a dataset from a real experiment.

### The problem

We investigate whether there is a relationship between the nonuse of the proximal part of the upper limb (PANU) and the nonuse of the shoulder (SANU) or the elbow (EENU).

## Methods

### Dataset

The dataset is far from perfect, which reflects the reality of experimental research, particularly in clinical contexts.

```{r load data NonUse}
nonuse <- read.csv("data/test_data/NonUse.csv", sep=",", header=TRUE)
head(nonuse)
summary(nonuse)
```
### Variables
This file contains measures of upper extremity nonuse:

    PANU: Proximal Arm Non Use
    SANU: Shoulder Antepulsion Non Use
    EENU: Elbow Extension Non Use


Description of the experiment, the variables and the data (units and the range of possible values).

| name | description | unit | range
|---|-------|--|-----
| PANU | Proximal Arm Non Use | % | O-100 |
| SANU | Shoulder Antepulsion Non Use | % | O-100 |
| EENU | Elbow Extension Non Use | % | O-100 |

## Analyses

### Corelation analysis

- Correlation between PANU/SANU

We used a Spearman correlation, because we don't have assumptions about the distribution of the data. We do not assume linear relationship between variables.
```{r correlation PANU-SANU}
cor.test(nonuse$PANU, nonuse$SANU, method = "spearman")
```
- Correlation between PANU/EENU
```{r correlation PANU-EENU}
cor.test(nonuse$PANU, nonuse$EENU, method = "spearman")
```

### PANU-SANU relationship

```{r plot SANU-lm}
library(ggplot2)
ggplot(nonuse, aes(x=PANU, y=SANU)) +
  geom_point() +
  geom_smooth(method='lm', formula= y~x, color='blue', se=T) +
  # xlim(-10, 60) +
  # ylim(-50, 50) +
  
labs(title="PANU vs. SANU", x="PANU (%)", y="SANU (%)")
```

### PANU-EENU relationship

```{r plot EENU-lm}

ggplot(nonuse, aes(x=PANU, y=EENU)) +
  geom_point() +
  geom_smooth(method='lm', formula= y~x, color='blue', se=T) +
  # xlim(-10, 60) +
  # ylim(-50, 50) +
  
labs(title="PANU vs. EENU", x="PANU (%)", y="EENU (%)")
```

### Linear regression for PANU-SANU

```{r lm/SANU}
lm_sanu <- lm(SANU ~ PANU, data=nonuse)
summary(lm_sanu)
coef(lm_sanu)
slope_1 <- coef(lm_sanu)[2]
intercept_1 <- coef(lm_sanu)[1] #?
cat("Regression equation: SANU=", slope_1, "PANU +", intercept_1, "\n")
```

### Linear regression for PANU-EENU

```{r lm/EENU}
lm_eenu <- lm(EENU ~ PANU, data=nonuse)
summary(lm_eenu)
coef(lm_eenu)
slope_2 <- coef(lm_eenu)[2]
intercept_2 <- coef(lm_eenu)[1] 
cat("Regression equation: EENU=", slope_2, "PANU +", intercept_2, "\n")
```
## Regression equations
(PANU = a * SANU + b and PANU = a * EENU + b)

```{r equations PANU}
a <- 1 / slope_1
b <- -intercept_1 / slope_1

cat("Regression equation : PANU =", a, "* SANU +", b, "\n")

# PANU = a*EENU + b
a <- 1 / slope_2
b <- -intercept_2 / slope_2

cat("Regression equation : PANU =", a, "* EENU +", b, "\n")
```

## Results

Spearman’s rank correlation rho showed a weak but significant positive association between PANU and SANU (ρ = 0.216, p = 0.0017), while a moderate and significant positive association was found between PANU and EENU (ρ = 0.387, p < 0.0001). Linear regression analyses confirmed these results: no significant relationship was observed between SANU and PANU (SANU = 0.089 × PANU + 0.586, R² = 0.009, p = 0.171), whereas a significant positive relationship was found between EENU and PANU (EENU = 0.565 × PANU + 3.284, R² = 0.225, p < 0.0001). When PANU was expressed as a function of SANU and EENU, the regression equations obtained were PANU = 11.195 × SANU – 6.561 and PANU = 1.771 × EENU – 5.816,  respectively.
(with GPT's help)

## Limits

The present analysis has several limitations.
First, the correlations observed were relatively weak, with rho values of 0.216 for PANU–SANU and 0.387 for PANU–EENU, indicating that only a small proportion of the variance is explained.
Second, the linear regression between PANU and SANU was not significant (R² = 0.009, p = 0.171), and even the significant regression between PANU and EENU explained only a modest proportion of variance (R² = 0.225).
Third, the dataset contains a high number of missing values (223 for SANU and EENU, 13 for PANU), reducing statistical power and potentially biasing the analysis.
In addition, the variables show large variability and extreme values, which may reflect outliers or measurement noise.
Finally, the use of linear regression assumes a linear relationship, which may not fully capture the complexity of the data
(with GPT's help)

# Statistical tests: comparison of medians (or means?)

In this series of exercises, the aim is to explore how to compare two groups of data, using non-parametric tests (or parametric tests).

## Definition

### Mean and median
The mean is the arithmetic average of a dataset. Median is the middle value that divides the dataset into two equal halves

- https://www.khanacademy.org/math/statistics-probability/summarizing-quantitative-data

### Variance and Standard Deviation (SD)
Variance measures the average squared distance from the mean, while the standard deviation is its square root, expressed in the same units as the data.

- https://www.statisticshowto.com/probability-and-statistics/statistics-definitions/variance/

### A normal distribution

A normal distribution is a symmetric, bell-shaped curve characterized by its mean and standard deviation.

- https://www.khanacademy.org/math/statistics-probability/modeling-distributions-of-data

### Statistical test
A statistical test is a method to decide if data provide enough evidence to reject a null hypothesis

- https://en.wikipedia.org/wiki/Statistical_hypothesis_testing

### When to use a statistical test?
A statistical test is used when we need to evaluate if an observed effect is unlikely to have occurred by chance.

- https://www.scribbr.com/statistics/statistical-tests/?utm_source=chatgpt.com

### Parametric test and non-parametric test
Parametric tests assume specific data distributions (e.g., normality), while non-parametric tests make fewer assumptions and work with ranks or ordinal data.

- https://www.scribbr.com/statistics/statistical-tests/?utm_source=chatgpt.com

### Assumptions of parametric tests
Parametric tests typically assume that the data (or residuals) are normally distributed, that variances are homogeneous across groups, that observations are independent, and that the variables are measured on an interval or ratio scale.

- https://www.scribbr.com/frequently-asked-questions/assumptions-of-statistical-tests/

### Assumptions of non-parametric tests
Non-parametric tests assume fewer conditions, mainly independence of observations and appropriate data scale (ordinal or rank).

- https://www.scribbr.com/statistics/nonparametric-tests/

### p-value
A p-value is the probability of observing results as extreme as those in your sample, assuming the null hypothesis is true.

- https://www.scribbr.com/statistics/p-value/

### What is the risk of error when using a statistical test?
A test involves risks: rejecting H0 when it is true (Type I error, α) or failing to reject H0 when H1 is true (Type II error, β).

- https://www.scribbr.com/statistics/type-i-and-type-ii-errors/

### What is the difference between a paired and an unpaired test?
A paired test compares related samples (e.g., before/after on the same subjects), while an unpaired test compares independent groups.

- https://www.scribbr.com/statistics/paired-t-test-vs-unpaired-t-test/


## Effect of treatment over time
In clinical and sport research, a common question is whether a treatment or training intervention produces an effect over time. To address this scientifically, measurements need to be collected before and after the intervention, and ideally at later time points, to evaluate both the immediate and the lasting effects

### The data
Download the file PrePost.csv
```{r load data PrePost}
prepost <- read.csv("data/test_data/PrePost.csv", sep=",", header=TRUE)
head(prepost)
summary(prepost)
```

This file contains before/after measurements. The treatment is a rehabilitation training for individuals with cardiac conditions.



| Variable | Description | Unit | Possible values | Expected change |

2.2.2 The analysis
You want to know if the treatment has an effect on one or more of the measured variables.

Does the treatment have an effect?

What comparison are you going to make?

Which test will you use? Parametric or non-parametric?

Which graph illustrates the effect of the treatment?

What sentence will you write in your thesis to explain your result?

2.3 Testing some stereotypes
Humans have stereotypes, some of them are probably true, some others are probably false (e.g., ref ).

Here, we will test stereotypes about snorers… among others.

2.3.1 The data
Download the file snore.txt

This file contains anthropometric and qualitative measurements (1 person per line).

2.3.2 The analysis
Do the data confirm the following stereotypes? Provide a reasoned answer (data, figure, result sentence) for each question.
Are snorers fatter?
Do snorers drink or smoke more?
Are men fatter?
Do women smoke less?
From a more general perspective:
Are there any correlations between variables?
